---
title: "Migrating old code to the new mRpostman's syntax"
output: 
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{Migrating old code to the new mRpostman's syntax}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

<style>
  .col2 {
    columns: 2 100px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 100px; /* chrome, safari */
    -moz-columns: 2 100px;    /* firefox */
  }
  <!-- .col3 { -->
  <!--   columns: 3 100px; -->
  <!--   -webkit-columns: 3 100px; -->
  <!--   -moz-columns: 3 100px; -->
  <!-- } -->
</style>

## Introduction

This vignette aims to provide an easy source of code migration examples so old `mRpostman` users can transition between the old version and the new format which implements an OO approach trough the use of R6 class and methods.

This is the summary of the main modifications in the package between version `0.3.1` and `0.9.0.0`:

* All main functions, except `list_attachments` and the custom-search helper functions, now are methods of the R6 class `ImapConf`;
    
* The way the connection token is passed through functions (now methods) has changed. The connection handle is created inside `configure_imap()` (or `ImapCon$new()`) and modified with custom requests from the methods that trigger IMAP commands. As a consequence, the password is now hidden inside the curl handle C pointer, resulting in a more secure token chain. This resulted in changes in every request function. These functions (that are methods now) do not use `config_handle()` anymore, and a call to `curl::set_opt()` is made in every request function so that a custom request is supplied or replaced by a new one in the original handle.
    
* The "by" argument used in search and fetch functions was replaced by `use_uid`, which is a logical now with the default value set as `FALSE`. This is equivalent to the former `by = MSN` default configuration.
    
* all functions that returned `invisible(0L)` now return `invisible(TRUE)`

* Users that prefer a tidy approach, only need to apply the pipe `%>%` operator between search and fetch methods (or other complementary operations), in which messages' ids are carried from one level to another. The exposition pipe `%$%` is not necessary anymore.


In the following sections, we present some code migration examples between the old and the new version of `mRpostman`.


## 1) Configuring the IMAP connection

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}

library(mRpostman)

# Outlook - Office 365
imapconf <- configure_imap(url="imaps://outlook.office365.com",
                           username="your_user@company.com",
                           password=rstudioapi::askForPassword())

# other mail providers that were tested: Gmail (imaps://imap.gmail.com), 
#   Hotmail ("imaps://imap-mail.outlook.com"), Yahoo (imaps://imap.mail.yahoo.com/), 
#   AOL (imaps://export.imap.aol.com/), Yandex (imaps://imap.yandex.com)

```
  
<br>
  
New format:
```{r, message = FALSE, eval=FALSE}

library(mRpostman)

# Outlook - Office 365
con <- configure_imap(url="imaps://outlook.office365.com",
                      username="your_user@company.com",
                      password=rstudioapi::askForPassword())

# alternative
con <- ImapCon$new(url="imaps://outlook.office365.com",
                   username="your_user@company.com",
                   password=rstudioapi::askForPassword())

```

</div>

Although the object that is created is different (now we have an R6 `ImapCon` class), the code to create a connection object is pretty much the same, but with an alternative that uses the "initialize" method of the R6 class.

Other useful options are `timeout_ms`, `verbose = TRUE`, and `buffersize`. Further 'curl' options related to IMAP functionalities can be passed to `configure_imap()`, depending on the libcurl version installed on the user's machine. See `curl::curl_options()`. In this new version, there is an important curl parameter exposed, `xoauth2_bearer`, which enables OAuth2.0 authentication. For more details, check the [_"IMAP OAuth2.0 authentication in mRpostman"_](https://allanvc.github.io/mRpostman/articles/xoauth2.0.html) vignette.


## 2) Listing Server Capabilities

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
imapconf %>%
  list_server_capabilities()
```

New format:
```{r, message = FALSE, eval=FALSE}
con$list_server_capabilities()
```

</div>


## 3) Mailbox Commands

### 3.1) Listing mail folders

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
# Listing
imapconf %>%
  list_mailboxes()
```

New format:
```{r, message = FALSE, eval=FALSE}
# Listing
con$list_folders()
```
</div>

### 3.2) Folder selection

<div class="col2">
Previous code:
```{r, message = FALSE, eval = FALSE}
imapconf %>%
  select_mailbox(mbox = "INBOX") # only INBOX is case sensitive
```

New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "INBOX")
```
</div>

### 3.3) Examining a mail folder

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  examine_mailbox()
```

New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "K-State")

con$examine_folder()
```

</div>

In the new format, the previously selected folder is kept during the session unless you execute a new selection.

### 3.4) Renaming a mail folder

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
imapconf %>%
  select_mailbox(mbox = "CRAN messages") %>%
  rename_mailbox(new_name = "CRAN messages2") %>%
  list_mailboxes() # and list again to check

```

<br>
<br>

New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "CRAN messages")

con$rename_folder(new_name = "CRAN messages2")

con$list_mail_folders() # and list again to check
```

</div>


## 4) Single Search

All search functions now return only the message ids.

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
result <- imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  search_before(date_char = "02-May-2019", 
                by = "UID", 
                flag = "UNANSWERED")

result$msg_id

```

<br>
<br>
<br>
<br>
New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "K-State")

result <- con$search_before(date_char = "02-May-2019",
                            use_uid = TRUE,
                            flag = "UNANSWERED")

result

```
</div>

You can also **NEGATE** the statement to search for messages **NOT BEFORE a date**:

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  search_before(date_char = "02-May-2019", 
                negate = TRUE, 
                by = "UID", 
                flag = "UNANSWERED",
                esearch = TRUE)

results$msg_id

```


New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "K-State")

result <- con$search_before(date_char = "02-May-2019",
                            use_uid = TRUE,
                            flag = "UNANSWERED",
                            negate = TRUE)

result

```
</div>

This pattern is applied to all SEARCH functions.

Special attention should be paid to the `search_string` and `search_flag` methods due to changes in the names and order of the arguments.

#### Search by string - example 1

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_string(section_or_field = "FROM", string = "hadley@rstudio.com",
                negate = TRUE) # not FROM "hadley@rstudio.com"

results$msg_id

```

<br>
<br>

New format:
```{r, message = FALSE, eval=FALSE}
con$select_folder(name = "INBOX")

result <- con$search_string(expr = "hadley@rstudio.com",
                            where = "FROM",
                            negate = TRUE) # not FROM "hadley@rstudio.com"

result

```
</div>

#### Search by string - example 2

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  search_string(section_or_field = "TEXT", string = "Dear Allan")

```

<br>
<br>

New format:
```{r, message = FALSE, eval=FALSE}
con$select_folder(name = "K-State")

result <- con$search_string(expr = "Dear Allan", where = "TEXT")

result

```

#### Search by flag

<div class="col2">
Previous code:
```{r, message = FALSE, eval = FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_flag(flag = "RECENT", by = "UID")

```

<br>
<br>
<br>

New format:
```{r, message = FALSE, eval=FALSE}
con$select_folder(name = "INBOX")

result <- con$search_flag(expr = "RECENT", use_uid = TRUE)

result

```
</div>

Now you can check which flags are available in a mail folder with the `list_flags()` method. Also, you can pass multiple flags as the search criterion.


## 5) Custom Search


<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  custom_search(custom_request = 
                 AND(
                   string(section_or_field = "FROM", string = "@toronto"),
                   before(date_char = "12-Apr-2019")
                   )
               )

results$msg_id

```


New format:
```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

result <- con$search(custom_request = 
                       AND(
                         string(expr = "@toronto", where = "FROM"),
                         before(date_char = "12-Apr-2019")
                         )
                     )
result

```

</div>


## 6) Fetch

### 6.1) Fetch body

<div class="col2">
Previous code:
```{r, message = FALSE, eval = FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  search_before(date_char = "10-Mar-2019", by = "UID") %$% #exposition pipe operator
  fetch_full_msg(imapconf = imapconf, msg_id = msg_id,
                 by="UID", write_to_disk = TRUE, keep_in_mem = TRUE,
                 partial = "0.789")

```

New format:
```{r, message = FALSE, eval = FALSE}
con$select_folder(name = "K-State")

results <- con$search_before(date_char = "10-Mar-2019", use_uid = TRUE) %>% #regular pipe
  con$fetch_body(use_uid = TRUE, write_to_disk = TRUE, 
                 keep_in_mem = TRUE, partial = "0.789")

```

</div>


### 6.2) Fetch header

<div class="col2">
Previous code:
```{r, message = FALSE, eval=FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "K-State") %>%
  search_since(date_char = "15-Aug-2019", by = "UID") %$% #exposition pipe operator
  fetch_msg_header(imapconf = imapconf, 
                   msg_id = msg_id, 
                   fields = c("DATE", "SUBJECT"),
                   by = "UID")

```

New format:
```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "K-State")

results <- con$search_since(date_char = "15-Aug-2019", use_uid = TRUE) %>% #regular pipe
  con$fetch_header(use_uid = TRUE, fields = c("DATE", "SUBJECT"))

```

</div>

### 6.3) Fetch text

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_since(date_char = "17-Aug-2019", by = "UID") %$% #exposition pipe operator
  fetch_msg_text(imapconf = imapconf, msg_id = msg_id, by = "UID", try_b64decode = TRUE)

```


New format:
```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

results <- con$search_since(date_char = "17-Aug-2019", use_uid = TRUE) %>% #regular pipe
  con$fetch_text(use_uid = TRUE, base64_decode = TRUE))

```

</div>

### 6.4) Fetch metadata

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_on(date_char = "10-May-2019", by = "UID") %$% #exposition pipe operator
  fetch_msg_metadata(imapconf, msg_id = msg_id, by = "UID",
                     metadata = c("INTERNALDATE", "UID", "ENVELOPE"))

```

New format:
```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

results <- con$search_since(date_char = "10-May-2019", use_uid = TRUE) %>% #regular pipe
  con$fetch_text(use_uid = TRUE, metadata = c("INTERNALDATE", "UID", "ENVELOPE"))

```

</div>


You can check metadata options with `metadata_options()`.


## 7) Attachments

### 7.1) Attachments listing

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}
imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_on(date_char = "23-Sep-2019") %$%
  fetch_full_msg(imapconf, msg_id=msg_id) %>%
  list_attachments()
```

<br>
<br>


New format:
```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

results <- con$search_on(date_char = "23-Sep-2019") %>% #regular pipe
  con$fetch_body() %>% # or fetch_text()
  list_attachments()

```

</div>

Note that `list_attachment()` is an independent function. It is not a method of the `ImapCon` class that depends on the connection object.

The new version brings an alternative fetching method (`fetch_atachments_list()`) for listing attachments without the need of previously executing a `fetch_body` or `fetch_text` operation:

```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

con$search_on(date_char = "23-Sep-2019") %>% #regular pipe
  con$fetch_attachments_list()

```


### 7.2) Attachments Extraction

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}
imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_on(date_char = "23-Sep-2019") %$%
  fetch_full_msg(imapconf, msg_id=msg_id) %>%
  get_attachments()

```

<br>
<br>

New format:
```{r, message = FALSE, eval=FALSE}
con$select_folder(name = "INBOX")

results <- con$search_on(date_char = "23-Sep-2019") %>% #regular pipe
  con$fetch_body() %>% # or fetch_text()
  get_attachments()

```

</div>

The new version brings an alternative fetching method (`fetch_atachments()`) for downloading attachments without the need of previously executing a `fetch_body` or `fetch_text` operation:

```{r, message = FALSE, eval=FALSE}

con$select_folder(name = "INBOX")

con$search_on(date_char = "23-Sep-2019") %>% #regular pipe
  con$fetch_attachments()

```



## 8) Complementary Operations

### 8.1) Copy message(s)

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "CRAN messages2") %>%
  search_since(date_char = "10-May-2019") %$% #exposition pipe operator
  copy_msg(imapconf = imapconf, msg_id = msg_id, to_mbox = "INBOX")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "CRAN messages2")

results <- con$search_since(date_char = "10-May-2019") %>%
  con$copy_msg(to_folder = "INBOX", reselect = FALSE)

```

</div>

### 8.2) Move message(s) (depends on MOVE capability)


<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "[Gmail]/Sent") %>%
  search_before(date_char = "10-may-2012") %$% #exposition pipe operator
  move_msg(imapconf = imapconf, msg_id = msg_id, to_mbox = "CRAN messages2")


```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "[Gmail]/Sent")

con$search_before(date_char = "10-May-2012") %>%
  con$move_msg(to_folder = "CRAN messages2", reselect = FALSE)

```

</div>



### 8.3) Search the minimum message ID with specific flag(s) (depends on ESEARCH capability)

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "[GMail]/Trash") %>%
  get_min_id(flag = "UNSEEN")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "[GMail]/Trash")

con$esearch_min_id(flag = "UNSEEN")

```

</div>


### 8.4) Search the maximum message ID with specific flag(s) (depends on ESEARCH capability)

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}
results <- imapconf %>%
  select_mailbox(mbox = "[Gmail]/Trash") %>%
  get_max_id(flag = "UNSEEN")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "[GMail]/Trash")

con$esearch_max_id(flag = "UNSEEN")

```

</div>

### 8.5) Delete message(s)

<div class="col2">
Previous code:


```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  delete_msg(msg_id = 66128)

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "INBOX")

con$delete_msg(msg_id = 66128)

```

</div>

### 8.6) Expunge

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "[Gmail]/Trash") %>%
  delete_msg(msg_id = 71772, by = "UID") %$%
  expunge(imapconf = imapconf, specific_UID = msg_id)

```

<br>
<br>
<br>

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "[GMail]/Trash")

con$delete_msg(msg_id = 71772, use_uid = TRUE) %>%
  con$expunge() # expunge the specific UID

# or, if the user wants to expunge the entire mail folder:
con$expunge()

```

</div>

### 8.7) Add/Remove/Replace flags

#### 8.7.1) Add flags

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_since(date_char = "18-Aug-2020", by = "UID") %$% #exposition pipe operator
  add_flags(imapconf = imapconf, msg_id = msg_id, flags_to_set = "\\Seen")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "INBOX")

con$search_since(date_char = "18-Aug-2020", use_uid = TRUE) %>%
  con$add_flags(use_uid = TRUE, flags_to_set = "\\Seen") # system flags need the "\\" prefix

```

</div>

#### 8.7.2) Remove flags

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_since(date_char = "18-Aug-2020", by = "UID") %$% #exposition pipe operator
  remove_flags(imapconf = imapconf, msg_id = msg_id, flags_to_unset = "\\Seen")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "INBOX")

con$search_since(date_char = "18-Aug-2020", use_uid = TRUE) %>%
  con$remove_flags(use_uid = TRUE, flags_to_unset = "\\Seen") # system flags need the "\\" prefix

```

</div>


#### 8.7.3) Replace flags

<div class="col2">
Previous code:

```{r, message = FALSE, eval = FALSE}

results <- imapconf %>%
  select_mailbox(mbox = "INBOX") %>%
  search_since(date_char = "18-Aug-2020", by = "UID") %$% #exposition pipe operator
  replace_flags(imapconf = imapconf, msg_id = msg_id, flags_to_set = "\\Seen")

```

New format:
```{r, message = FALSE, eval = FALSE}

con$select_folder(name = "INBOX")

con$search_since(date_char = "18-Aug-2020", use_uid = TRUE) %>%
  con$replace_flags(use_uid = TRUE, flags_to_set = "\\Seen") # system flags need the "\\" prefix

```

</div>


