---
title: "IMAP OAuth2.0 authentication in mRpostman"
output:
  rmarkdown::html_vignette:
    toc: yes
vignette: >
  %\VignetteIndexEntry{IMAP OAuth2.0 authentication in mRpostman}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction {#intro}

Most major mail providers have moved to OAuth2.0 as their preferred (and, in
many cases, only) authentication method for IMAP. Instead of sending your
password, you send a short-lived **access token** that authorizes a specific
scope of access to your mailbox.

This vignette shows the exact steps and R code to obtain, use, and refresh an
OAuth2.0 access token for the **Gmail** IMAP server. The same logic applies to
other providers (Outlook/Office 365, Yahoo, AOL, ...) by swapping the
authorization endpoints and scope. The reference for Google's flow is
<https://developers.google.com/identity/protocols/oauth2>.

We use the [`httr`](https://cran.r-project.org/package=httr) package to run the
OAuth2.0 flow and [`jsonlite`](https://cran.r-project.org/package=jsonlite) to
read the credentials file, so install them first if needed:

```{r, eval = FALSE}
install.packages(c("httr", "jsonlite"))
```

__IMPORTANT — libcurl version.__ The libcurl feature that transmits the bearer
token is only reliable on libcurl >= 7.65.0 (released 2019-05-22). Check the
version the `curl` R package is linked against:

```{r, message = FALSE}
curl::curl_version()$version
```

If it is older, update libcurl (and reinstall `curl`) before continuing,
otherwise you may get a `SASL` error during authentication.

The whole process is:

1. create a Google Cloud project;
2. configure the OAuth consent screen (scope + test user);
3. create a **Desktop app** OAuth client and download its JSON credentials;
4. obtain an access token in R;
5. open the IMAP connection with `configure_imap(xoauth2_bearer = ...)`;
6. refresh the token when it expires.

Steps 1--3 are done once, in the browser; steps 4--6 are R code.

## Step 1 - Create a Google Cloud project

1. Go to <https://console.cloud.google.com/> and sign in with the Google
   account whose mailbox you want to access.
2. In the project selector at the top, click **New Project**, give it a name
   (e.g. `mRpostman`), and click **Create**.
3. Make sure this new project is the one selected in the top bar.

You do **not** need to enable the Gmail API: IMAP access is granted by the
OAuth scope `https://mail.google.com/`, not by the Gmail API. IMAP is enabled by
default on Gmail accounts.

## Step 2 - Configure the OAuth consent screen

Navigate to **APIs & Services > OAuth consent screen** (in the redesigned
console this section is called **Google Auth Platform**).

1. **User type:** choose **External** and click **Create**.
2. Fill in the required fields: **App name**, **User support email**, and, at
   the bottom, a **Developer contact email**. Save and continue.
3. **Scopes:** click **Add or remove scopes**, then, in the *"manually add
   scopes"* box, paste:

   ```
   https://mail.google.com/
   ```

   Click **Add to table**, then **Update** and **Save and continue**. (This is a
   *restricted* scope; while your app stays in **Testing** mode it works for
   test users without Google verification.)
4. **Test users:** add the e-mail address of the account you will log in with
   (e.g. `your_user@gmail.com`) and save.

The account you authenticate with in Step 4 **must** be listed here as a test
user, otherwise Google returns an `access_denied` error.

## Step 3 - Create the OAuth client credentials

1. Go to **APIs & Services > Credentials**.
2. Click **+ Create credentials > OAuth client ID**.
3. For **Application type**, select **Desktop app** (this is important --- a
   *Web application* client will fail with `redirect_uri_mismatch` in the flow
   below).
4. Give it a name and click **Create**.
5. Click **Download JSON** and save the file locally. Keep it private --- it
   contains your client secret.

The downloaded file looks like this (an `"installed"` client):

```json
{"installed":{"client_id":"XXXX.apps.googleusercontent.com",
  "project_id":"your-project","auth_uri":"https://accounts.google.com/o/oauth2/auth",
  "token_uri":"https://oauth2.googleapis.com/token",
  "client_secret":"GOCSPX-XXXX","redirect_uris":["http://localhost"]}}
```

Read the client id and secret from it in R:

```{r, eval = FALSE}
cred <- jsonlite::fromJSON("path/to/client_secret_XXXX.json")$installed
```

## Step 4 - Obtain an access token

Below are two ways to get a token. **Option A** is the shortest and works out of
the box on most desktops. Use **Option B** if Option A fails to start its local
web server (common inside RStudio, or on headless/remote machines).

Both request `access_type = "offline"` so that a **refresh token** is also
returned (see Step 6).

### Option A - `httr`'s built-in flow (recommended)

`httr::oauth2.0_token()` opens your browser, runs a local callback server, and
caches the token in a `.httr-oauth` file.

```{r, eval = FALSE}
library(httr)

gmail_app <- oauth_app("mRpostman",
                       key    = cred$client_id,
                       secret = cred$client_secret)

token_obj <- oauth2.0_token(
  endpoint = oauth_endpoints("google"),
  app      = gmail_app,
  scope    = "https://mail.google.com/",
  cache    = TRUE
)

token <- token_obj$credentials$access_token
```

When the browser shows *"Google hasn't verified this app"*, click
**Advanced > Go to \<your app\> (unsafe)** --- this is expected while the app is
in Testing mode --- then **Allow**.

### Option B - Manual loopback flow (no local server)

This avoids the local callback server entirely: you open the authorization URL,
approve access, and paste back the `code` shown in the browser's address bar.

```{r, eval = FALSE}
library(httr)

redirect_uri <- "http://localhost"
scope        <- "https://mail.google.com/"

auth_url <- modify_url("https://accounts.google.com/o/oauth2/auth",
  query = list(client_id     = cred$client_id,
               redirect_uri  = redirect_uri,
               response_type = "code",
               scope         = scope,
               access_type   = "offline",
               prompt        = "consent"))

browseURL(auth_url)
```

After you approve, the browser is redirected to
`http://localhost/?code=...&scope=...`. The page will fail to load ("unable to
connect") --- that is fine. Copy the value between `code=` and `&scope` from the
**address bar** and paste it below (authorization codes are single-use and
expire in minutes, so exchange it right away):

```{r, eval = FALSE}
code_in <- "PASTE_THE_CODE_HERE"

resp <- POST("https://oauth2.googleapis.com/token", encode = "form",
  body = list(code          = code_in,
              client_id     = cred$client_id,
              client_secret = cred$client_secret,
              redirect_uri  = redirect_uri,
              grant_type    = "authorization_code"))

token_data <- content(resp)
token      <- token_data$access_token
```

Your `token` should be a long string starting with `"ya29."`. If `token` is
`NULL`, inspect `content(resp)` --- an `invalid_grant` error means the code was
already used or expired, so re-run `browseURL(auth_url)` to get a fresh one.

## Step 5 - Open the IMAP connection

Pass the access token to `configure_imap()` via the `xoauth2_bearer` argument
(no `password`):

```{r, eval = FALSE}
library(mRpostman)

con <- configure_imap(
  url            = "imaps://imap.gmail.com",
  username       = "your_user@gmail.com",
  use_ssl        = TRUE,
  xoauth2_bearer = token
)

con$list_server_capabilities()
```

If this returns the server's capabilities, you are authenticated and can use any
`mRpostman` method (`select_folder()`, `search_*()`, `fetch_*()`, ...).

## Step 6 - Refresh the token

Access tokens are short-lived (about 1 hour). The **refresh token** returned in
Step 4 (thanks to `access_type = "offline"`) lets you mint a new access token
without going through the browser again.

If you used **Option A**, `httr` refreshes automatically; you can also force it:

```{r, eval = FALSE}
token_obj$refresh()
token <- token_obj$credentials$access_token
```

If you used **Option B**, exchange the stored refresh token directly:

```{r, eval = FALSE}
refresh <- POST("https://oauth2.googleapis.com/token", encode = "form",
  body = list(client_id     = cred$client_id,
              client_secret = cred$client_secret,
              grant_type    = "refresh_token",
              refresh_token = token_data$refresh_token))

token <- content(refresh)$access_token
```

Then reopen (or update) the connection with the new `token`. To swap the token
on an existing connection object, use `con$reset_xoauth2_bearer(token)`.

## Troubleshooting

- **`access_denied` / "hasn't been verified by Google":** the account you logged
  in with is not in the consent screen's **Test users** list (Step 2.4), or it
  is not the account that owns the project. Add it as a test user.
- **`redirect_uri_mismatch`:** your OAuth client is a *Web application*, not a
  *Desktop app*. Recreate it as a **Desktop app** (Step 3).
- **`createTcpServer: address already in use` (Option A):** the local callback
  port is busy --- often a leftover server from a previous attempt in the same
  R session. Run `httpuv::stopAllServers()` and retry; if that fails, fully quit
  and reopen RStudio (a mere "Restart R" does not release the port), or switch to
  **Option B**, which needs no local server.
- **`SASL` error during authentication:** your libcurl is likely older than
  7.65.0. Update it and reinstall the `curl` R package (see the
  [Introduction](#intro)).
- **Other providers:** replace the authorization/token endpoints and the scope
  with the provider's values (for Google, `oauth_endpoints("google")` already
  fills them in), and set `url` accordingly, e.g.
  `imaps://outlook.office365.com` for Office 365.
